Natural selection in the Health and Retirement Study

### 2023-09-20 hrs-selection.qmd
Don't use dead siblings (N reduced too much). Report agefb negative correlation in scores.
Give up on furrr. Figure out bug in confint.svrepdesign. Save time reading rand_fam. Skip many regressions in black sample.

confint_workaround <- function (x) {
  cbind(
    as.numeric(x) - 1.96 * SE(x), 
    as.numeric(x) + 1.96 * SE(x))
}

@@

# rand family respondents file
rand_fam <- haven::read_dta(
  "data/randhrsfam1992_2018v2_STATA/randhrsfamr1992_2018v2.dta")
# only selected columns to avoid creating suffixes when we join with
# rand.
# Age of oldest kid; grandchildren; num siblings who died
rand_fam <- rand_fam |> select(hhid, pn, h10ageokid, h10gkid, r10sbdied)

@@

rand <- rand |>
  mutate(.by = rabyear, # calculate this pooling ethnicity
    mean_raevbrn = mean(raevbrn, na.rm = TRUE),
    rlrs = raevbrn/mean_raevbrn
  ) |>
  mutate(.by = ethnicity, # calculate these stats within-ethnicity
    income_resid = calc_income(rand, ethnicity[1]),
    income_resid = income_resid - min(income_resid, na.rm = TRUE) + 1,
    income_med = chop_equally(income_resid, 2, labels = c("Low", "High")),
    # as.numeric avoids haven.labelled issues:
    raedyrs = as.numeric(raedyrs),
    edu = chop(as.numeric(raedyrs), 13, lbl_discrete(symbol = "-")),
    # wave 10 is 2010
    # table(rands$white$r10mstat)
     # 1                 1.married
     # 2   2.married,spouse absent
     # 3               3.partnered
     # 4               4.separated
     # 5                5.divorced
     # 6      6.separated/divorced
     # 7                 7.widowed
     # 8           8.never married
    married = factor(r10mstat == 1, 
                     levels = c(FALSE, TRUE),
                     labels = c("Other", "Married")),
    r10sibs = r10livsib + r10sbdied + 1, # n. sibs including oneself!
    agefb = r10agey_e - h10ageokid,
    agefb = ifelse(agefb < 12, NA_real_, agefb),
    agefb = chop_equally(agefb, 2, labels = c("Low", "High"))
  )

@@

  form_pgs <- map(pgs, \(x) reformulate(c(x, pcs), response = "rlrs"))
  mod_pgs[[eth]] <- map(form_pgs, 
                        \(x) svyglm_quiet(x, r_eth)
                        )

@@

  form_pgs_edyrs <- map(pgs, \(x) reformulate(c(x, pcs),
                                              response = "raedyrs"))
  mod_pgs_edyrs[[eth]] <- map(form_pgs_edyrs, 
                              \(x) svyglm_quiet(x, r_eth)
                              )

@@

  # the colon interaction estimates the effect separately within each group
  form_pgs_x_ed <- paste0(pgs, ":edu") |>
    map(\(x) reformulate(c(x, "edu", pcs), response = "rlrs"))
  mod_pgs_x_ed[[eth]]  <-
    map(form_pgs_x_ed, 
        \(x) svyglm_quiet(x, r_eth)
        )

@@

  form_pgs_x_inc <- paste0(pgs, ":income_med") |>
    map(\(x) reformulate(c(x, "income_med", pcs), response = "rlrs"))
  mod_pgs_x_inc[[eth]]  <-
    map(form_pgs_x_inc, 
        \(x) svyglm_quiet(x, r_eth)
        )

@@

  form_pgs_x_mar <- paste0(pgs, ":married") |>
    map(\(x) reformulate(c(x, "married", pcs), response = "rlrs"))
  mod_pgs_x_mar[[eth]]  <-
    map(form_pgs_x_mar, 
        \(x) svyglm_quiet(x, r_eth)
        )

@@

  form_pgs_x_agefb <- paste0(pgs, ":agefb") |>
    map(\(x) reformulate(c(x, "agefb", pcs), response = "rlrs"))
  mod_pgs_x_agefb[[eth]]  <-
    map(form_pgs_x_agefb, 
        \(x) svyglm_quiet(x, r_eth)
        )

  tidy_pgs_x_agefb[[eth]] <-
    map(mod_pgs_x_agefb[[eth]], broom::tidy, conf.int = TRUE) |> 
    setNames(pgs) |> 
    list_rbind(names_to = "pgs") |>
    filter(! term %in% c("(Intercept)", "agefbHigh", pcs)) |>
    mutate(
      agefb = str_remove(term, ".*:agefb")
    )

tidy_pgs_eth <- bind_rows(tidy_pgs,  .id = "ethnicity")

n_tests <- nrow(tidy_pgs_eth)
n_sig_bonf <- sum(tidy_pgs_eth$p.value < 0.05/n_tests)

@@

  cor(coef_pgs_black, coef_pgs_white)
}

cor_eths <- cor(tidy_pgs$white$estimate, tidy_pgs$black$estimate)
cor_eths_boot <- withReplicates(rand_boot, theta = calc_eth_cor)
ci_eths <- confint_workaround(cor_eths_boot)[1,]
```

@@

```{r}
#| label: stat-bootstraps
#| cache: true
#| cache-lazy: false

n_reps <- 599

@@

  c(
    cor(coef_pgs_edu, coef_pgs_fert),
    coef(lm(coef_pgs_fert ~ coef_pgs_edu))["coef_pgs_edu"]
  )
}

@@

@@

  # this is positive if white was "bigger" than black taking the white sign 
  # as the correct sign. Negative if b was "bigger"  than w.
  diff_coefs <- (coef_pgs_white - coef_pgs_black)*sign(coef_pgs_white)
  names(diff_coefs) <- pgs
  mean(diff_coefs)
}

@@

ci_boot_fert <- map(cors_fert_edu, \(x) confint_workaround(x)[1,])
ci_boot_sibs <- map(cors_sibs_edu, \(x) confint_workaround(x)[1,])

ci_boot_fert_lm <- map(cors_fert_edu, \(x) confint_workaround(x)[2,])
ci_boot_sibs_lm <- map(cors_sibs_edu, \(x) confint_workaround(x)[2,])

stopifnot(significant(ci_boot_fert$white))
stopifnot(insignificant(ci_boot_fert$black))
stopifnot(insignificant(ci_boot_sibs$white))
stopifnot(insignificant(ci_boot_sibs$black))
```

@@

I can also examine natural selection in the previous generation, by regressing PGS on respondents' number of siblings (including dead siblings) in 2010. I reweight respondents by the reciprocal of their number of siblings, to account for parents of many siblings being more likely to be a parent of a respondent. Parents of no siblings cannot be included, so coefficient sizes are not comparable across the generations. Appendix @fig-siblings plots coefficents on number of siblings versus coefficients on years of education. Correlations are insignificant for both ethnic groups, although note that standard errors are large (whites: correlation `r cors_sibs_edu$white[1]`, bootstrapped 95% C.I. `r ci_boot_sibs$white[1]` to `r ci_boot_sibs$white[2]`; blacks: correlation `r cors_sibs_edu$black[1]`, bootstrapped 95% C.I. `r ci_boot_sibs$black[1]` to `r ci_boot_sibs$black[2]`).

@@

@@

  for (term in terms) {
    diff_cor_ed[[eth]][[term]] <- 
      withReplicates(rand_boot, theta = calc_diff_cor_ed, term = term)  
    ci[[eth]][[term]] <- confint_workaround(diff_cor_ed[[eth]][[term]])
    est[[eth]][[term]] <- as.numeric(diff_cor_ed[[eth]][[term]])
  }
}

# TODO: does splitting by education to 12 and education 13-16 mechanically
#   cause a lower correlation with edyrs in the 13-16 group?

@@

```{r}
#| label: stat-x-bootstrap-effect-size
#| cache: true
#| cache-lazy: false

n_reps_x <- 99

@@

rand_boot <- rand |> 
  subset(ethnicity == "white") |> 
  as.svrepdesign(type = "bootstrap", replicates = n_reps_x)

for (term in terms) {
  diff_effect_size[[term]] <- 
    withReplicates(rand_boot, theta = calc_diff_effect_size, term = term)  
  ci[[term]] <- confint_workaround(diff_effect_size[[term]])
  est[[term]] <- as.numeric(diff_effect_size[[term]])
  tbl[[term]] <- sprintf("%.3f (%.3f to %.3f)", est[[term]], 
                         ci[[term]][, 1], ci[[term]][, 2])
}

@@

stopifnot(insignificant(ci$edu[3,]))
stopifnot(significant(ci$income_med[3,]))
stopifnot(significant(ci$married[3,]))
stopifnot(insignificant(ci$agefb[3,]))
```

Appendix @fig-x-edu shows essentially no difference in the distribution of absolute coefficient sizes between respondents with more and less education. Appendix @fig-x-inc, on the other hand, shows that coefficients appear larger among lower-income respondents. Appendix @fig-x-mar shows absolute coefficients for respondents who were married in 2010 against all other statuses. Coefficients appear larger for unmarried respondents. Appendix @fig-x-agefb splits respondents by age at first birth (median within each ethnicity). There are no large differences between early and late parents. Note that the N is lower here due to missing values. @tbl-x-bootstraps shows means and 95% confidence intervals for mean absolute coefficient sizes in the two groups, and the difference between them, using bootstraps. Differences for income and marriage are significant.

@@

: Bootstrap estimates and 95% confidence intervals for mean absolute coefficients of PGS on RLRS in "low" and "high" groups among white respondents. Groups are: 0-12 years education vs. 13-17 years; below vs. above median income; all others vs. married; below vs. above median age at first birth. `r n_reps_x` bootstraps. {#tbl-x-bootstraps tbl-colwidths="[16,28,28,28]"}

>>>>

@@

# rand family respondents file
# Done just once:
# rand_fam <- haven::read_dta(
#   "data/randhrsfam1992_2018v2_STATA/randhrsfamr1992_2018v2.dta")
# saveRDS(rand_fam, "data/randhrsfam1992_2018v2_STATA/rand_fam.rds")
rand_fam <- readRDS("data/randhrsfam1992_2018v2_STATA/rand_fam.rds")

# only selected columns to avoid creating suffixes when we join with
# rand.
# Age of oldest kid; grandchildren; num siblings who died
rand_fam <- rand_fam |> select(hhid, pn, h10ageokid, h10gkid, r10sbdied)

@@

rand <- rand |>
  mutate(.by = rabyear, # calculate this pooling ethnicity
    mean_raevbrn = mean(raevbrn, na.rm = TRUE),
    rlrs = raevbrn/mean_raevbrn
  ) |>
  mutate(.by = ethnicity, # calculate these stats within-ethnicity
    income_resid = calc_income(rand, ethnicity[1]),
    income_resid = income_resid - min(income_resid, na.rm = TRUE) + 1,
    income_med = chop_equally(income_resid, 2, labels = c("Low", "High")),
    # as.numeric avoids haven.labelled issues:
    raedyrs = as.numeric(raedyrs),
    edu = chop(as.numeric(raedyrs), 13, lbl_discrete(symbol = "-")),
    # wave 10 is 2010
    # table(rands$white$r10mstat)
     # 1                 1.married
     # 2   2.married,spouse absent
     # 3               3.partnered
     # 4               4.separated
     # 5                5.divorced
     # 6      6.separated/divorced
     # 7                 7.widowed
     # 8           8.never married
    married = factor(r10mstat == 1, 
                     levels = c(FALSE, TRUE),
                     labels = c("Other", "Married")),
    r10sibs = r10livsib + 1, # n. sibs including oneself!
    agefb = r10agey_e - h10ageokid,
    agefb = ifelse(agefb < 12, NA_real_, agefb),
    agefb = chop_equally(agefb, 2, labels = c("Low", "High"))
  )

@@

  form_pgs <- map(pgs, \(x) reformulate(c(x, pcs), response = "rlrs"))
  mod_pgs[[eth]] <- map(form_pgs, 
                               \(x) svyglm_quiet(x, r_eth))

@@

  form_pgs_edyrs <- map(pgs, \(x) reformulate(c(x, pcs),
                                              response = "raedyrs"))
  mod_pgs_edyrs[[eth]] <- map(form_pgs_edyrs, 
                                     \(x) svyglm_quiet(x, r_eth))

@@

# we only look at interactions among whites, the rest is pointless
for (eth in "white") {
  r_eth <- rand |> subset(ethnicity == eth)

  # the colon interaction estimates the effect separately within each group
  form_pgs_x_ed <- paste0(pgs, ":edu") |>
    map(\(x) reformulate(c(x, "edu", pcs), response = "rlrs"))
  mod_pgs_x_ed[[eth]]  <-
    map(form_pgs_x_ed, 
               \(x) svyglm_quiet(x, r_eth))

@@

  form_pgs_x_inc <- paste0(pgs, ":income_med") |>
    map(\(x) reformulate(c(x, "income_med", pcs), response = "rlrs"))
  mod_pgs_x_inc[[eth]]  <-
    map(form_pgs_x_inc, 
               \(x) svyglm_quiet(x, r_eth))

@@

  form_pgs_x_mar <- paste0(pgs, ":married") |>
    map(\(x) reformulate(c(x, "married", pcs), response = "rlrs"))
  mod_pgs_x_mar[[eth]]  <-
    map(form_pgs_x_mar, 
               \(x) svyglm_quiet(x, r_eth))

@@

  form_pgs_x_agefb <- paste0(pgs, ":agefb") |>
    map(\(x) reformulate(c(x, "agefb", pcs), response = "rlrs"))
  mod_pgs_x_agefb[[eth]]  <-
    map(form_pgs_x_agefb, 
               \(x) svyglm_quiet(x, r_eth))

  tidy_pgs_x_agefb[[eth]] <-
    map(mod_pgs_x_agefb[[eth]], broom::tidy, conf.int = TRUE) |> 
    setNames(pgs) |> 
    list_rbind(names_to = "pgs") |>
    filter(! term %in% c("(Intercept)", "agefbHigh", pcs)) |>
    mutate(
      agefb = str_remove(term, ".*:agefb")
    ) 
}

tidy_pgs_eth <- list_rbind(tidy_pgs)
n_tests <- nrow(tidy_pgs_eth)
n_sig_bonf <- sum(tidy_pgs_eth$p.value < 0.05/n_tests)

@@

  c(cor = cor(coef_pgs_black, coef_pgs_white))
}

cor_eths <- cor(tidy_pgs$white$estimate, tidy_pgs$black$estimate)
cor_eths_boot <- withReplicates(rand_boot, theta = calc_eth_cor)
ci_eths <- confint(cor_eths_boot)[1,]
```

@@

```{r}
#| label: stat-bootstraps
#| cache: true
#| cache-lazy: false
#| cache-comments: false

# Not sure this actually does anything :-)
old_opts <- options(survey.multicore = TRUE)

# The large n_reps is to get an accurate upper tail on the black sample, 
# since it is close to 0:
n_reps <- 599

@@

  c(
    cor = cor(coef_pgs_edu, coef_pgs_fert),
    coef = coef(lm(coef_pgs_fert ~ coef_pgs_edu))["coef_pgs_edu"]
  )
}

@@

options(old_opts)
```

@@

  # this is positive if white was "bigger" than black taking the white sign 
  # as the correct sign. Negative if b was "bigger"  than w.
  diff_coefs <- (coef_pgs_white - coef_pgs_black)*sign(coef_pgs_white)
  names(diff_coefs) <- pgs
  c(diff = mean(diff_coefs))
}

@@

ci_boot_fert <- map(cors_fert_edu, \(x) confint(x)[1,])
ci_boot_sibs <- map(cors_sibs_edu, \(x) confint(x)[1,])

ci_boot_fert_lm <- map(cors_fert_edu, \(x) confint(x)[2,])
ci_boot_sibs_lm <- map(cors_sibs_edu, \(x) confint(x)[2,])

if (insignificant(ci_boot_fert$white)) warning("ci_boot_fert$white insignificant")
if (! insignificant(ci_boot_fert$black)) warning("ci_boot_fert$black significant")
if (significant(ci_boot_sibs$white)) warning("ci_boot_sibs$white significant")
if (significant(ci_boot_sibs$black)) warning("ci_boot_sibs$black significant")
```

@@

I can also examine natural selection in the previous generation, by regressing PGS on respondents' number of living siblings in 2010. (Data for dead siblings has too many missing values to use.) I reweight respondents by the reciprocal of their number of siblings, to account for parents of many siblings being more likely to be a parent of a respondent. Parents of no siblings cannot be included, so coefficient sizes are not comparable across the generations. Appendix @fig-siblings plots coefficients on number of siblings versus coefficients on years of education. Correlations are insignificant for both ethnic groups, although note that standard errors are large (whites: correlation `r cors_sibs_edu$white[1]`, bootstrapped 95% C.I. `r ci_boot_sibs$white[1]` to `r ci_boot_sibs$white[2]`; blacks: correlation `r cors_sibs_edu$black[1]`, bootstrapped 95% C.I. `r ci_boot_sibs$black[1]` to `r ci_boot_sibs$black[2]`).

@@

old_opts <- options(survey.multicore = TRUE)

@@

  for (term in terms) {
    diff_cor_ed[[eth]][[term]] <- 
      withReplicates(rand_boot, theta = calc_diff_cor_ed, term = term)  
    ci[[eth]][[term]] <- confint(diff_cor_ed[[eth]][[term]])
    est[[eth]][[term]] <- as.numeric(diff_cor_ed[[eth]][[term]])
  }
}

options(old_opts)

@@

```{r}
#| label: stat-x-bootstrap-effect-size
#| cache: true
#| cache-lazy: false
#| cache-comments: false

old_opts <- options(survey.multicore = TRUE)

n_reps_x_es <- 99

@@

rand_boot <- rand |> 
  subset(ethnicity == "white") |> 
  as.svrepdesign(type = "bootstrap", replicates = n_reps_x_es)

for (term in terms) {
  diff_effect_size[[term]] <- 
    withReplicates(rand_boot, theta = calc_diff_effect_size, term = term)  
  ci[[term]] <- confint(diff_effect_size[[term]])
  est[[term]] <- as.numeric(diff_effect_size[[term]])
  tbl[[term]] <- sprintf("%.3f (%.3f to %.3f)", est[[term]], 
                         ci[[term]][, 1], ci[[term]][, 2])
}

options(old_opts)

```{r}
#| label: stat-agefb-bootstrap

n_reps_agefb <- 9
calc_cor_agefb <- function (w, data) {
  forms <- pgs |> map(\(x) reformulate(c(x, pcs), response = "raevbrn"))
  mods_low <- map(forms, \(x) lm(x, data, subset = agefb == "Low",
                                 weights = w))
  mods_high <- map(forms, \(x) lm(x, data, subset = agefb == "High",
                                  weights = w))
  coefs_low <- map2_dbl(mods_low, pgs, \(x, y) coef(x)[[y]])
  coefs_high <- map2_dbl(mods_high, pgs, \(x, y) coef(x)[[y]])

  c(
    cor = cor(coefs_low, coefs_high), 
    discordant = sum(sign(coefs_low) != sign(coefs_high)) # discordant pairs
  )
}

rand_boot <- rand |> 
  subset(ethnicity == "white") |> 
  as.svrepdesign(type = "bootstrap", replicates = n_reps_agefb)

cor_agefb <- withReplicates(rand_boot, theta = calc_cor_agefb)

stopifnot(insignificant(confint(cor_agefb)[1,]))
ci_discordant <- confint(cor_agefb)[2,]
# significantly different from 1/2?
stopifnot(ci_discordant[1] <= length(pgs)/2 && 
            ci_discordant[2] >= length(pgs)/2)

@@

stopifnot(insignificant(ci$edu[3,]))
stopifnot(significant(ci$income_med[3,]))
stopifnot(significant(ci$married[3,]))
stopifnot(insignificant(ci$agefb[3,]))

tidy_agefb_wide <- tidy_pgs_x_agefb$white |> 
  pivot_wider(id_cols = pgs, names_from = agefb, values_from = estimate)

cor_agefb <- cor(tidy_agefb_wide$Low, tidy_agefb_wide$High)
n_discordant <- sum(sign(tidy_agefb_wide$Low) != sign(tidy_agefb_wide$High))
```

Appendix @fig-x-edu shows essentially no difference in the distribution of absolute coefficient sizes between respondents with more and less education. Appendix @fig-x-inc, on the other hand, shows that coefficients appear larger among lower-income respondents. Appendix @fig-x-mar shows absolute coefficients for respondents who were married in 2010 against all other statuses. Coefficients appear larger for unmarried respondents.

Appendix @fig-x-agefb splits respondents by age at first birth (median within each ethnicity). There are no large differences between younger and older parents. As in the UK Biobank sample, correlations of PGS coefficients between younger and older parents are negative (`r cor_agefb`), and for `r n_discordant` out of `r length(pgs)` PGS, coefficients have opposite signs. However, in bootstraps, these statistics are not significantly different from 0 and `r length(pgs)/2` respectively. Note that the N is lower here due to missing values.

@tbl-x-bootstraps shows means and 95% confidence intervals for mean absolute coefficient sizes in each pair of groups, and the difference between them, using bootstraps. Differences for income and marriage are significant.

@@

: Bootstrap estimates and 95% confidence intervals for mean absolute coefficients of PGS on RLRS in "low" and "high" groups among white respondents. Groups are: 0-12 years education vs. 13-17 years; below vs. above median income; all others vs. married; below vs. above median age at first birth. `r n_reps_x_es` bootstraps. {#tbl-x-bootstraps tbl-colwidths="[16,28,28,28]"}
